An enhancer code compendium of early human neuronal and neural crest development

### 2025-04-09 manuscript.tex
Changes throughout text.

\affil[1]{VIB Center of AI \& computational Biology, Leuven, Belgium}
\affil[2]{VIB-KU Leuven Center for Brain and Disease Research, Leuven, Belgium}
\affil[3]{KU Leuven Department of Human Genetics, Leuven, Belgium}
\affil[4]{Karolinska Insitute, Department of Biochemistry and Biophysics, Stockholm, Sweden}
\affil[5]{Karolinska Institutet, Division of Neurodegeneration, Department of Neurobiology, Care Sciences and Society,Stockholm, Sweden}

@@

%Context
Enhancers encode cell types as clusters of transcription factor binding sites. 
%need
Studies of enhancer-codes are predominantly limited to a small number of cell types at a time and often focus
on model organisms.
%task and results
In this study we generate a single-cell multiome atlas of differentiating human neural tube organoids 
and of the head of a four weeks post conception human embryo. 
Using DNA sequence based deep learning - DeepNeuralTube - we identify a total of 15 cell types that 
are in common between the organoid model and the human embryo and we provide a compendium of enhancer-codes 
for each individual cell type. 
We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 transcription factor binding sites in a single nucleosome depleted region. 
Enhancers encoding anterior-posterior patterning, however, form a exception to this rule where they are encoded using
homotypic clusters of homeodomain transcription factor binding sites.
%conclusion
%implication

@@

Genomes of complex multi-cellular organisms specify all cell types and states\cite{trapnellDefiningCellTypes2015, arendtOriginEvolutionCell2016, fleckWhatCellType2023} through distinct non-coding elements named
enhancers\cite{davidsonRegulatoryGenomeGene2010}. 
Central to the study of the genomic code of enhancers is the paradigm that 
enhancers consist of heterotypic clusters of transcription factor binding sites (TFBS)\cite{longEverChangingLandscapesTranscriptional2016,kimDecipheringMultiscaleQuantitative2023} with soft syntactic rules\cite{zeitlingerSevenMythsHow2020} 
and that cooperative binding of multiple transcription factors (TFs)\cite{lambertHumanTranscriptionFactors2018} underlies cell type-specific activity of the enhancer.\par

DNA sequence based deep learning methods\cite{
  alipanahiPredictingSequenceSpecificities2015,
  zhouPredictingEffectsNoncoding2015,
  kelleyBassetLearningRegulatory2016,
  quangDanQHybridConvolutional2016,
  movvaDecipheringRegulatoryDNA2019,
  kelleySequentialRegulatoryActivity2018,
  yuanScBassetSequencebasedModeling2022,
  chenSequencebasedGlobalMap2022,
  maslovaDeepLearningImmune2020,
  minnoyeCrossspeciesAnalysisEnhancer2020,
  avsecBaseresolutionModelsTranscriptionfactor2021,
  avsecEffectiveGeneExpression2021,
  dealmeidaDeepSTARRPredictsEnhancer2022,
  linderPredictingRNAseqCoverage2025a,
  kempynckCREstedModelingGenomic2025} have shown major advances in the ability to model and explain enhancers\cite{zeitlingerSevenMythsHow2020,dewinterModellingDesignTranscriptional2025} 
and largely support the enhancer-code paradigm. 
However, these studies are predominantly limited to a small number of cell type or cell states\cite{
  maslovaDeepLearningImmune2020,
  minnoyeCrossspeciesAnalysisEnhancer2020,
  atakInterpretationAllelespecificChromatin2021,
  avsecEffectiveGeneExpression2021,
  bravogonzalez-blasSinglecellSpatialMultiomics2024}, and often focus on model organisms\cite{
  janssensDecodingGeneRegulation2022,
  heckerEnhancerdrivenCellType2025}, with only few studies profiling human development\cite{mannensChromatinAccessibilityHuman2024}.\par

@@

Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols 
and how they compare to their \textit{in vivo} counterparts at a transcriptomic level\cite{
  heIntegratedTranscriptomicCell2024,
  xuIntegratedTranscriptomicCell2023}. 
However, such comparisons at the level of the enhancer-code are still limited\cite{
  zenkSinglecellEpigenomicReconstruction2024,
  fleckInferringPerturbingCell2023}.

In this study, we address these limitations by building an enhancer-codes compendium of cell types and cell states generated by human neural tube organoids, 
a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}.
We compare this compendium to that of the head of a four weeks post conception human embryo.
Through these comparisons we confirm the enhancer-code paradigm, 
but also provide a counter examples specifically related to the enhancer-code underlying anterior-posterior patterning.
\par

@@

To evaluate the biological relevance of the cell types and states obtained from the organoid, 
we also performed 10x single cell multiome on the head of a single four weeks post conception human embryo, 
resulting in 25,785 high quality cells (\textbf{Fig. 1c}). 
We annotated both the organoid and embryo cells based on the expression of marker genes
(\textbf{Supplementary Table S1}). 
In both systems we could identify neuronal progenitos, early differentiating neurons,
neural crest, facial mesenchyme and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) 
while perivascular macrophages (PVM) / microglia (MIC) and otic placode cells were missing (\textbf{Fig. 1c}).\par

Next, we sub-clustered the neuronal progenitor, early neuron, and neural crest cell types (\textbf{Fig. 1d-f}). 
Both organoid and embryo neuronal progenitors show a gradient of different cell states reflecting dorso-ventral patterning, evident from the expression of dorso-ventral marker TFs\cite{delasRepressiveInteractionsGene2020}:
\textit{ARX}, \textit{NKX2-2}, \textit{PAX6}, \textit{IRX3}, \textit{ZIC1} and \textit{GDF7} (\textbf{Fig. 1d}). 
In both the organoid and embryo system we obtained multiple early neuronal differentiation states.
This includes a cluster that expresses the neuronal progenitor marker \textit{SOX2}, cells that express the early neuronal induction marker \textit{ASCL1} and distinct clusters marked by the expression of \textit{GAD1}, \textit{UNCX} and \textit{GATA2} (\textbf{Fig. 1e}). 
Finally, focusing on the neural crest, in both the organoid and the embryo we identified a population of pre-migratory neural crest (\textit{GRHL2} expression and absence of \textit{ZEB2} expression), migratory neural crest (\textit{SOX10} and \textit{ZEB2} expression), facial mesenchyme (\textit{TWIST1} expression) and early differentiation peripheral neurons (\textit{NEUROD1} expression). 
In the embryo, but not in the organoid, we observed an additional population of cells marked by the expression of \textit{SOX10} and \textit{PAX2} reflecting the otic placode (\textbf{Fig. 1f}).\par

Next, we assessed the expression of anterior-posterior marker TFs. 
While the organoid cells had either an anterior forebrain(\textit{SIX3}) or anterior spinal cord (\textit{HOXB3}) identity, 
in the embryo we observed a gradient of anterior-to-posterior identities. Marked by the expression of \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1 }and \textit{HOXB3} (\textbf{Fig. 1g})\cite{braunComprehensiveCellAtlas2023}.\par

@@

Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo on a chromatin accessibility level 
we identified sets of co-accessible regions in the organoid model using topic modeling. 
Focusing on topics that were enriched for marker TF motifs we calculated the average expression of marker genes for each cell group where these regions were co-accessible for both organoid and embryo cells. 
We observed similar patterns of marker gene expression for the same cell state (defined by
the topic) across the organoid and embryo (\textbf{Fig. 1i}) 
with an average Pearson correlation coefficient of 0.83 across identical states (\textbf{Supplementary Fig. S1c-d}).\par

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_1.png}
  \caption{
    \textbf{Human neural tube organoids recapitulate early neuronal and neural crest development.} \\
    \textbf{a}, Schematic. \textbf{b-c}, UMAP based on gene expression profiles of organoid (\textbf{b}) and embryo
    (\textbf{c}) cells colored by cell type identify. \textbf{d-f}, UMAP based on gene expression profiles of organoid (top)
    and embryo (bottom) cells types showing sub-clusters of neuronal progenitors (\textbf{d}), early differentiating neurons
    (\textbf{e}) and neural crest (\textbf{f}) along with expression of indicated genes in red-green-blue color scale.
    (Continues next page)
  }
\end{figure}
\afterpage{\clearpage}

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) \textbf{g}, UMAP based on gene expression profiles of organoid (top; UMAP axis 1 and 2 shown) and embryo (bottom; UMAP
    axis 2 and 3 shown) for neuronal progenitor subcluster along with the expression of indicated genes in red-green-blue
    color scale. \textbf{h}, UMAP based on gene expression profiles in organoid (top left) and embryo (top right) colored by
    cell cycle phase identity and quantification of percentage of each cell for each cell cycle phase stratified by cell
    type in the organoid (bottom left) and embryo (bottom right). \textbf{i}, Average gene expression level in
    $log_{2}(CPM)$
    (dot color) and fraction of cells expressing the gene (dot size) per cell state, defined by topic modeling, for organoid
    cells (top) and embryo cells (middle) and motif enrichment level as normalized enrichment score (NES; dot color) and
    number of regions enriched for the motif (dot size) per topic (x-axis) and motif (y-axis; bottom). PVM/MIC: perivascular
    microphage/microglia.
    Color scales for RGB scatter plots are as follows (in Log(CPM)). 
    For organoid: 
    Minimum is 0 and maximum is 1.5 Log(CPM) for all genes except:
    IRX3, BAHRL1: [0, 1.0], 
    FOXG1, GAD1, UNXC, GATA2, SOX10: [0, 0.8] and 
    ZEB2 [0, 3.0], 
    For embryo: 
    Minimum is 0 and maximum is 1.5 Log(CPM) for all genes except:
    FOXG1, GAD1, UNCX, GATA2, SOX10 [0, 0.8], 
    ISL1 [0, 1.0], 
    ZEB2 [0, 3.0] and
    TWIST1 [0, 2.0].
  }
\end{figure}

@@

In order to gain insight into the sequence characteristics underlying cell type-specific activity of neural tube-, 
neural crest- and derived cell type enhancers, we trained two sequence-to-function model convolutional neural networks:
"$DeepNeuralTube^o$" and "$DeepNeuralTube^e$". 
The models use DNA sequence of genomic regions as input and classify regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class fashion; \textbf(Fig. 2a)). 
The classes of the models are defined based on topic modeling. 
To get a fine-grained definition of cell states, we performed topic modeling separately for the subset of neuronal progenitors, neural crest (incl. facial mesenchyme and peripheral neurons) and early differentiating neurons for both organoid (to train $DeepNeuralTube^o$) embryo (to train $DeepNeuralTube^e$).\par

We tested the performance of $DeepNeuralTube^o$ and $DeepNeuralTube^e$ to predict cell type-specific activity 
using data from the VISTA enhancer browser\cite{viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}.
Enhancers active in the neural tube (n=349) could be classified with an area under the receiver-operator curve
(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2b}; and
area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Supplementary Fig. S2a-b})). 
Enhancers active in the facial mesenchyme (n=143) could be classified with an auROC of 0.78 and 0.79 
using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2c}; 
and auPR of 0.17 for both models (\textbf{Supplementary Fig. S2a-b})). 
Of note, a fraction of neural tube enhancers (24 \%) are predicted to be specific to pre-migratory neural crest by
both models. Given that the VISTA enhancers are annotated based on localization of enhancer activity only, and given
that is is difficult to visually distinguish between the neural tube and pre-migratory neural crest, this result 
is not surprising.\par

Next, we assessed the explainability of the models. Concretely we tested if the model learned known
TF binding motifs as features. To this end, we calculated pairwise similarities\cite{
  guptaQuantifyingSimilarityMotifs2007} between \textit{de novo} motifs learned by the model, 
and identified using TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018}, and known motifs that are enriched in cell type specific regions.
Clustering this pairwise similarity matrix resulted in 56 clusters (\textbf{Fig. 2d}). 
Most clusters contain known motifs enriched in regions specific to both organoid and embryo cell types (\textbf{Fig. 2e}), with the exception of cluster 13 that is organoid specific (AP1 motif; \textbf{Supplementary Table S2}) 
and cluster 26 that is embryo specific (CG-rich motif; \textbf{Supplementary Table S2}). 
More importantly, Most clusters contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Supplementary Table S2}). 
%TODO: add percentages
Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based
on cell type-specific chromatin accessibility.\par

As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{Supplementary Fig. S2c}). 
This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). 
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ explain this region similarly (Pearson correlation of x) 
%TODO: Add correlation
and highlight the importance of multiple SOX TFBS (\textbf{Fig. 2h}). 
A version of this enhancer where 5\% of the nucleotides are changed (hs111.1) does not have neural tube specific enhancer activity and both models predict that this loss of activity is due to the loss of a sinlge SOX TFBS (\textbf{Fig. 2h}).\par

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_2.png}
  \caption{
    \textbf{DeepNeuralTube: an accurate sequence-to-function model revealing neuro-developmental enhancer code.} \\
    \textbf{a}, Schematic. \textbf{b} receiver-operator curves for identifying neural tube (left) and facial mesenchyme the
    $DeepNeuralTube^o$ and $DeepNeuralTube^e$. \textbf{c-e}, UMAP of TomTom [REF] motif-to-motif similarities
    using both \textit{de novo} motifs from DeepNeuralTube and known enriched motifs. Each dot is a motif. Motifs are
    colored by: cluster identity (\textbf{c}), model of origin (Continues next page)
  }
\end{figure}
\afterpage{\clearpage}

\begin{figure}[b]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) (\textbf{d}; only showing \textit{de novo} motifs in this case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
 \textbf{f}, Example representative motifs from clusters in b, showing motifs from motif enrichment (left) and found
    \textit{de novo} (right). \textbf{g}, contribution score for reference and mutant version of Vista enhancer hs111
    for $DeepNeuralTube^o$ (top two panels) and $DeepNeuralTube^e$ (bottom two panels) and enhancer-reporter
    assay for reference and mutant version of hs111 (right two panels) [REF]. \textbf{h}, contribution score for
    $DeepNeuralTube^o$ (top) and $DeepNeuralTube^e$ (bottom) of a genomic enhancer human neural crest
    enhancer (chr1:161200798-161201298; hg38) and enhancer-reporter assay in chicken embryo (HH4; right). Green is enhancer signal, pink is
    constitutive enhancer that serves as electroporation control.
  }
\end{figure}

@@

Focusing on neural crest cells, we identified topics representing cells and regions of both pre- and migratory neural
crest as well as facial mesenchyme (\textbf{Fig. 5a}). Similar to the neuronal progenitors, we used
$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube^e$ to generate a neural crest code-table (\textbf{Fig. 5b}). 
Again, both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns and the importance of the instances of
these patterns is highly correlated across both models (0.73 average Pearson correlation coefficient; \textbf{Fig. 5b}).\par

@@

Focusing on early neuronal differentiation, we identified several topics for different stages of neuronal fate
acquisition in both organoid and embryo cells (\textbf{Fig. 6a}). Again we made use of $DeepNeuralTube^o$ and
$DeepNeuralTube^e$ to generate a neuron code-table (\textbf{Fig. 6b}). 
Both models found similar patterns and the importance of the instances of these patterns is highly correlated across
both models (0.47 average Pearson correlation coefficient; \textbf{Fig. 6b}).\par

>>>>

\affil[1]{VIB Center of AI \& computational Biology, Leuven, Belgium}
\affil[2]{VIB-KU Leuven Center for Brain and Disease Research, Leuven, Belgium}
\affil[3]{KU Leuven Department of Human Genetics, Leuven, Belgium}
\affil[4]{Karolinska Institute, Department of Biochemistry and Biophysics, Stockholm, Sweden}
\affil[5]{Karolinska Institute, Division of Neurodegeneration, Department of Neurobiology, Care Sciences and Society,Stockholm, Sweden}

@@

%Context
Enhancers encode cell types as clusters of transcription factor binding sites. 
%need
Enhancer code studies are predominantly limited to a small number of cell types often focusing
on model organisms.
%task and results
In this study we generate a single-cell multiome atlas of human neural tube organoids 
and of the head of a four weeks post conception human embryo. 
Using DNA sequence based deep learning - DeepNeuralTube - we identify a total of 15 cell types that 
are shared between the organoid model and the human embryo and we provide a compendium of enhancer codes 
for each individual cell type. 
We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 transcription factor binding sites in a nucleosome depleted region. 
Enhancers encoding anterior-posterior patterning form an exception to this rule where they are encoded using
homotypic clusters of homeodomain binding sites.
%conclusion
This compendium of enhancer codes thus provides gene regulatory insight into human development.
%implication

@@

Genomes of complex multi-cellular organisms specify all cell types and states\cite{trapnellDefiningCellTypes2015, arendtOriginEvolutionCell2016, fleckWhatCellType2023} through distinct non-coding elements named
enhancers\cite{davidsonRegulatoryGenomeGene2010}. 
Central to the study of the genomic code is the paradigm that 
enhancers consist of heterotypic clusters of transcription factor binding sites (TFBS)\cite{longEverChangingLandscapesTranscriptional2016,kimDecipheringMultiscaleQuantitative2023} with soft syntactic rules\cite{zeitlingerSevenMythsHow2020} 
and that cooperative binding of multiple transcription factors (TFs)\cite{lambertHumanTranscriptionFactors2018} underlies cell type-specific activity of the enhancer.\par

DNA sequence based deep learning methods\cite{
  alipanahiPredictingSequenceSpecificities2015,
  zhouPredictingEffectsNoncoding2015,
  kelleyBassetLearningRegulatory2016,
  quangDanQHybridConvolutional2016,
  movvaDecipheringRegulatoryDNA2019,
  kelleySequentialRegulatoryActivity2018,
  yuanScBassetSequencebasedModeling2022,
  chenSequencebasedGlobalMap2022,
  maslovaDeepLearningImmune2020,
  minnoyeCrossspeciesAnalysisEnhancer2020,
  avsecBaseresolutionModelsTranscriptionfactor2021,
  avsecEffectiveGeneExpression2021,
  dealmeidaDeepSTARRPredictsEnhancer2022,
  linderPredictingRNAseqCoverage2025a,
  kempynckCREstedModelingGenomic2025} have shown major advances in the ability to model and explain enhancers\cite{zeitlingerSevenMythsHow2020,dewinterModellingDesignTranscriptional2025} 
and largely support the enhancer code paradigm. 
However, these studies are predominantly limited to a small number of cell type or cell states\cite{
  maslovaDeepLearningImmune2020,
  minnoyeCrossspeciesAnalysisEnhancer2020,
  atakInterpretationAllelespecificChromatin2021,
  avsecEffectiveGeneExpression2021,
  bravogonzalez-blasSinglecellSpatialMultiomics2024}, and often focus on model organisms\cite{
  janssensDecodingGeneRegulation2022,
  heckerEnhancerdrivenCellType2025}, with only few studies profiling human development\cite{mannensChromatinAccessibilityHuman2024}.\par

@@

Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols 
and how they compare to their \textit{in vivo} counterparts at a transcriptomic level\cite{
  heIntegratedTranscriptomicCell2024,
  xuIntegratedTranscriptomicCell2023}. 
However, such comparisons at the level of the enhancer code are still limited\cite{
  zenkSinglecellEpigenomicReconstruction2024,
  fleckInferringPerturbingCell2023}.

In this study, we address these limitations by building an enhancer code compendium of cell types and cell states generated by human neural tube organoids, 
a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}.
We compare this compendium to that of the head of a four weeks post conception human embryo.
Through these comparisons we confirm the enhancer code paradigm, 
but also provide a counter examples specifically related to the enhancer code underlying anterior-posterior patterning.
\par

@@

To evaluate the biological relevance of the cell types and states obtained from the organoid, 
we also performed 10x single cell multiome on the head of a single four weeks post conception human embryo, 
resulting in 25,785 high quality cells (\textbf{Fig. 1c}). 
We annotated both the organoid and embryo cells based on the expression of marker genes
(\textbf{Supplementary Table S1}). 
In both systems we could identify neuronal progenitors, early differentiating neurons,
neural crest, facial mesenchyme and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) 
while perivascular macrophages (PVM) / microglia (MIC) and otic placode cells were missing (\textbf{Fig. 1c}).\par

Next, we sub-clustered the neuronal progenitor, early neuron, and neural crest cells (\textbf{Fig. 1d-f}). 
Both organoid and embryo neuronal progenitors show a gradient of different cell states reflecting dorso-ventral patterning, evident from the expression of dorso-ventral marker TFs\cite{delasRepressiveInteractionsGene2020}:
\textit{ARX}, \textit{NKX2-2}, \textit{PAX6}, \textit{IRX3}, \textit{ZIC1} and \textit{GDF7} (\textbf{Fig. 1d}). 
In both the organoid and embryo system we obtained multiple early neuronal differentiation states.
This includes a cluster that expresses the neuronal progenitor marker \textit{SOX2}, cells that express the early neuronal induction marker \textit{ASCL1} and distinct clusters marked by the expression of \textit{GAD1}, \textit{UNCX} and \textit{GATA2} (\textbf{Fig. 1e}). 
Finally, focusing on the neural crest, in both the organoid and the embryo we identified a population of pre-migratory neural crest (\textit{GRHL2} expression and absence of \textit{ZEB2} expression), migratory neural crest (\textit{SOX10} and \textit{ZEB2} expression), facial mesenchyme (\textit{TWIST1} expression) and early differentiation peripheral neurons (\textit{NEUROD1} expression). 
In the embryo, but not in the organoid, we observed an additional population of cells marked by the expression of \textit{SOX10} and \textit{PAX2} reflecting the otic placode (\textbf{Fig. 1f}).\par

Next, we assessed the expression of anterior-posterior (AP) marker TFs. 
While the organoid cells had either an anterior forebrain (\textit{SIX3}) or anterior spinal cord (\textit{HOXB3}) identity, 
in the embryo we observed a gradient of anterior-to-posterior identities marked by the expression of \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1 }and \textit{HOXB3} (\textbf{Fig. 1g})\cite{braunComprehensiveCellAtlas2023}.\par

@@

Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo on a chromatin accessibility level 
we identified sets of co-accessible regions in the organoid model using topic modeling 
and identified equivalent cells in the embryo where these same regions were also co-accessible.
We observed similar patterns of marker gene expression for the same cell state across the organoid and embryo (\textbf{Fig. 1i}) 
with an average Pearson correlation coefficient of 0.83 across identical states (\textbf{Supplementary Fig. S1c-d}).\par

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_1.png}
  \caption{
    \textbf{Human neural tube organoids recapitulate early neuronal and neural crest development.} \\
    \textbf{a}, Schematic of organoid differentiating protocol and human embryo experiment 
    \textbf{b-c}, UMAP based on gene expression profiles of organoid (\textbf{b}) and embryo
    (\textbf{c}) cells colored by cell type identify. 
    (Continues next page)
  }
\end{figure}
\afterpage{\clearpage}

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued)
    \textbf{d-f}, UMAP based on gene expression profiles of organoid (top) and embryo (bottom) cells types showing sub-clusters of neuronal progenitors (\textbf{d}), early differentiating neurons (\textbf{e}) and neural crest (\textbf{f}) along with expression of indicated genes in red-green-blue color scale.
    \textbf{g}, UMAP based on gene expression profiles of organoid (top; UMAP axis 1 and 2 shown) and embryo (bottom; UMAP
    axis 2 and 3 shown) for neuronal progenitor subcluster along with the expression of indicated genes in red-green-blue
    color scale. 
    \textbf{h}, UMAP based on gene expression profiles in organoid (top left) and embryo (top right) colored by
    cell cycle phase identity and quantification of percentage of each cell cycle phase stratified by cell
    type in the organoid (bottom left) and embryo (bottom right). 
    \textbf{i}, Average gene expression level in $log_{2}(CPM)$ (dot color) and fraction of cells expressing the gene (dot size) per cell state, defined by topic modeling, for organoid
    cells (top) and embryo cells (middle) and motif enrichment level as normalized enrichment score (NES; dot color) and
    number of regions enriched for the motif (dot size) per topic (x-axis) and motif (y-axis; bottom). PVM/MIC: perivascular
    microphage/microglia.
    Color scales for RGB scatter plots are as follows (in Log(CPM)). 
    For organoid: 
    Minimum is 0 and maximum is 1.5 for all genes except:
    IRX3, BAHRL1: [0, 1.0], 
    FOXG1, GAD1, UNXC, GATA2, SOX10: [0, 0.8] and 
    ZEB2 [0, 3.0], 
    For embryo: 
    Minimum is 0 and maximum is 1.5 for all genes except:
    FOXG1, GAD1, UNCX, GATA2, SOX10 [0, 0.8], 
    ISL1 [0, 1.0], 
    ZEB2 [0, 3.0] and
    TWIST1 [0, 2.0].
  }
\end{figure}

@@

In order to gain insight into the sequence characteristics underlying cell type-specific chromatin accessibility of 
neuronal progenitor, neural crest and neuronal cell type-specific enhancers, 
we trained two sequence-to-function model convolutional neural networks:
"$DeepNeuralTube^o$" and "$DeepNeuralTube^e$". 
The models use the DNA sequence of genomic regions as input and classify regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class fashion; \textbf{Fig. 2a}). 
The classes of the models are defined based on topic modeling. 
To get a fine-grained definition of cell states, we performed topic modeling separately for the subset of neuronal progenitors, neural crest (incl. facial mesenchyme and peripheral neurons) and early differentiating neurons for both organoid (to train $DeepNeuralTube^o$) embryo (to train $DeepNeuralTube^e$).\par

We tested the performance of $DeepNeuralTube^o$ and $DeepNeuralTube^e$ to predict cell type-specific activity 
using data from the VISTA enhancer browser\cite{viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}.
Enhancers active in the neural tube (n=349) could be classified with an area under the receiver-operator curve
(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2b}; and
area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Supplementary Fig. S2a-b})). 
Enhancers active in the facial mesenchyme (n=143) could be classified with an auROC of 0.78 and 0.79 
using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2c}; 
and auPR of 0.17 for both models (\textbf{Supplementary Fig. S2a-b})). 
Of note, a fraction of neural tube enhancers (\SI{24}{\percent}) are predicted to be specific to pre-migratory neural crest by
both models. Given that the VISTA enhancers are annotated based on localization of enhancer activity only, and given
that it is difficult to visually distinguish between the neural tube and pre-migratory neural crest, this result 
is not surprising.\par

Next, we assessed the explainability of the models. Concretely we tested if the model learned known
TF binding motifs as features. To this end, we calculated pairwise similarities\cite{
  guptaQuantifyingSimilarityMotifs2007} between \textit{de novo} motifs learned by the model, 
and identified using TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018}, and known motifs that are enriched in cell type specific regions.
Clustering this pairwise similarity matrix resulted in 56 clusters (\textbf{Fig. 2d}). 
All clusters, except cluster 47 (organoid specific CG motif), contain known motifs enriched in regions specific to both 
organoid and embryo cell types (\textbf{Fig. 2e} and \textbf{Supplementary Table S2}).
More importantly, Most clusters (\SI{70}{\percent}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Supplementary Table S2}). 
Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based
on cell type-specific chromatin accessibility.\par

As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{Supplementary Fig. S2c}). 
This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). 
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ explain this region similarly (Pearson correlation of 0.79) 
and highlight the importance of multiple SOX TFBS (\textbf{Fig. 2h}). 
A version of this enhancer where 5\% of the nucleotides are changed (hs111.1) does not have neural tube specific enhancer activity and both models predict that this loss of activity is due to the loss of a sinlge SOX TFBS (\textbf{Fig. 2h}).\par

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_2.png}
  \caption{
    \textbf{DeepNeuralTube: an accurate sequence-to-function model revealing neuro-developmental enhancer code.} \\
    \textbf{a}, Precision recall curves for $DeepNeuralTube^o$ and $DeepNeuralTube^e$ colored by main class and 
    average area under (AUC) the precision recall curves per class (left) and schematic representation of the model
    (right)
    (Continues next page)
  }
\end{figure}
\afterpage{\clearpage}

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) 
    \textbf{b}, receiver-operator curves for classifying neural tube (left) and facial mesenchyme enhancers using
    $DeepNeuralTube^o$ and $DeepNeuralTube^e$. 
    \textbf{c-e}, UMAP of TomTom\cite{guptaQuantifyingSimilarityMotifs2007} motif-to-motif similarities
    using both \textit{de novo} motifs from DeepNeuralTube and known enriched motifs. Each dot is a motif. Motifs are
    colored by: cluster identity (\textbf{c}), model of origin 
    (\textbf{d}; only showing \textit{de novo} motifs in this case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
    \textbf{f}, Example representative motifs from clusters in b, showing motifs from motif enrichment (left) and found
    \textit{de novo} (right). 
    \textbf{g}, contribution score for reference and mutant version of VISTA enhancer hs111
    for $DeepNeuralTube^o$ (top two panels) and $DeepNeuralTube^e$ (bottom two panels) and enhancer-reporter
    assay for reference and mutant version of hs111 (right two panels)\cite{
    viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}. 
    \textbf{h}, contribution score for
    $DeepNeuralTube^o$ (top) and $DeepNeuralTube^e$ (bottom) of a genomic human neural crest
    enhancer (chr1:161200798-161201298; hg38) and enhancer-reporter assay in chicken embryo (HH4; right). Green is enhancer signal, pink is
    constitutive enhancer that serves as electroporation control.
  }
\end{figure}

@@

Focusing on neural crest cells, we identified topics representing cells and regions of both pre- and migratory neural
crest as well as facial mesenchyme (\textbf{Fig. 5a}). Similar to the neuronal progenitors, we used
$DeepNeuralTube^o$ and $DeepNeuralTube^e$ to generate a neural crest code-table (\textbf{Fig. 5b}). 
Again, both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns and the importance of the instances of
these patterns is highly correlated across both models (0.73 average Pearson correlation coefficient; \textbf{Fig. 5b}).\par

@@

Focusing on early neuronal differentiation, we identified several topics for different stages of neuronal fate
acquisition in both organoid and embryo cells (\textbf{Fig. 6a}). Again we made use of $DeepNeuralTube^o$ and
$DeepNeuralTube^e$ to generate a neuron code-table (\textbf{Fig. 6b}). 
Both models found similar patterns and the importance of the instances of these patterns is correlated across
both models (0.47 average Pearson correlation coefficient; \textbf{Fig. 6b}).\par
